Titeln på avhandlingens språk

### 2023-01-23 publications/2021ImpactDelayedResponse/body.tex
still fixing paper c

\cref{item:pervasive} implies use of lightweight and low-power devices, preferably a wearable device that frees both hands for work.
\cref{item:interaction}, on the other hand, suggests a level of context sensitivity and proactivity that can only be provided by real-time analysis of sensor inputs such as video and audio feeds.
The compute-intensive processing suggested by \cref{item:interaction} cannot be met by the lightweight wearable devices suggested by \cref{item:pervasive}.
Only by offloading computation from a wearable device to cloud-based or edge-based infrastructure can this circle be squared.
However, offloading implies an extended end-to-end pipeline with many potential sources of queueing, transmission, and processing delays.
\cref{item:lowlatency} therefore emerges as a key concern, requiring deep understanding of the impact of end-to-end delays on \gls{WCA} users.

@@

The structure of this paper is as follows.
We describe the existing body of research around time perception and the effects of delay on human performance in \cref{sec:background}.
\cref{sec:experimentaldesign} presents the experimental design, measures, and specific protocol.
Then, in \cref{sec:results} we detail the results of our experiment.
Implications for further modeling of the effects of delay are presented in \cref{sec:discussion} before finally concluding the paper in \cref{sec:conclusion}.

\section{Background and Related Work}\label{sec:background}

@@

\subsection{Time perception in computing systems}\label{ssec:timeperception}

@@

\subsection{Mechanisms Relating Delay to Human Performance}\label{ssec:potentialmechs}

@@

\section{Experimental Design}\label{sec:experimentaldesign}

The core elements of our experiment are shown in \cref{fig:experimentaltestbed}:
\begin{itemize}
    \item Subjects interact with a \gls{WCA} while wearing an array of biometric sensors.
    \item The \emph{responsiveness} of the application, i.e.\ the interval of time between an input being provided to the system and the associated output returned to the user, is manipulated in real time. 
    The effects of these manipulations on the subjects are recorded and subsequently analyzed.    
\end{itemize}

\begin{figure}[h]
  \centering
  \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig1.eps}
  \caption{Experimental test-bed.
  Participants interact with the cognitive assistant through task-related inputs and outputs --- in practice, these correspond to the video feed captured by the assistant and the instructions provided by it.
  The assistant itself has been instrumented with a data collection layer, which collects and processes experiment-related data such as biometric signals from the participants (these are merely processed here and do not form part of the inputs to the cognitive assistant as such, however), and a delay buffer, which introduces controlled delays in the transit of information from the core processing component.}%
  \label{fig:experimentaltestbed}
\end{figure}

@@

% In more formal terms, we can provide definitions for \emph{task}, \emph{subtask}, and \emph{step} in such an application as:
In the following, we will provide some formal definitions relating to the \gls{WCA} application.
% \theoremstyle{definition}
% \begin{definition}[Task and subtask]
A \emph{task} will be understood as a finite sequence of instructions to be performed in order.
\emph{Step} will refer to a specific action to be performed by the user, described by a single instruction; e.g. \emph{put a green \(1 \times 3\) brick on the board} (see \cref{fig:task:steps}).
% We will refer to the interval of time delimited by two consecutive instructions (see \cref{fig:task:steps,fig:cogassist:step}), characterized by the actions of both the user and the assistant.
A step starts with an instruction being given to the user, and ends when the assistant detects that the user has finished performing it.
At that point, the assistant either provides a new instruction to the user, or the task ends.
  % See for instance \cref{fig:task:steps}, which pictures the task of assembling a simple LEGO model, with each subtask corresponding to the addition of a specific LEGO piece to the current model.  
% \end{definition}

\begin{figure*}[h]
  \centering
  \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig2.eps}
  \caption{Example of a Cognitive Assistance task and its component steps.}\label{fig:task:steps}
\end{figure*}

@@

\begin{figure*}[h]
  \centering
  % \input{plots/lego_hierarchy_diagram.tex}
  \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig3.eps}
  \caption{Hierarchical cognitive structure of a step in the LEGO task.}
  \label{fig:lego:hierarchical}
\end{figure*}

@@

For the experimental LEGO task, a key modification was made to the structure of the steps.
After each the processing of each input is completed, the result is withheld for a variable period of time until a specific target \emph{delay} is reached, as illustrated in \cref{fig:cogassist:step,fig:cogassist:step:delay}.
The length of this delay is one of two independent variables we manipulated for the experiment.
Seven levels of delay were used --- no added delay (which we will also refer to as \SI{0}{\second} delay), \SIlist{0.6;1.125;1.65;2.175;2.7;3.0}{\second} --- chosen based on the latency bounds found previously by Chen et al.~\cite{Chen:AnEmpiricalStudyOfLatency}.
% A value of  \SI{600}{\milli\second} was identified as the bound where users start noticing delays in the assistant.
% Conversely, \SI{2.7}{\second} was identified as the upper bound on delays after which the application is considered to be in such a degraded state it is basically ``unusable''.
Our selection of delays is centered around the range of delays where latency is noticeable to users, but the application remains in a ``usable'' state.
% while including one delay value in the unnoticeable range and one completely in the ``unusable'' range.

In order to study the effects of a delay applied across multiple steps we implement an experimental design component called a \emph{block}, corresponding to a sequence of consecutive steps within a task subject to the same delay; see \cref{fig:cogassist:block}.
% \begin{definition}[Block]
%   A sequence of consecutive steps within a task subject to the same delay; see \cref{fig:cogassist:block}.
The \emph{length} of blocks is the second of the two independent variables manipulated in our experiment.
We used values of \numlist{4;8;12} steps, chosen as representative of the number of steps in tasks in the base LEGO Assistant application.
Additionally, we define the \emph{duration} of a block, as the time elapsed between the start timestamp of the first step in the block and the end timestamp of the final step in the block.
% ; e.g.\ for \cref{fig:cogassist:block}, the duration of the pictured block would be \( t_{n+k} - t_{n} \).
% \end{definition}

@@

For an arbitrary step, we define the sequence \( \{ t_{0}, t_{1}, \ldots, t_{n} \} \) as the sequence of timestamps corresponding to the sampling instants during the step (see \cref{fig:exectime:diagram}).
\( t_{0} \) corresponds to the instant when the instruction for the step is given to the user (and the first sample is taken), and \( t_{n} \) to the timestamp of the last sample before a new instruction is given.
If we define \( t_{c} \) as the instant marking when the user presented the finished step to the system, it must follow that \( t_{n-1} < t_{c} < t_{n} \).
Due to the discrete sampling of the task state, there remains some imprecision in the estimate of execution time relative to \( t_{c} \).
However, this introduces no bias in the results, as we can assume that \( t_{c} \) is uniformly distributed in the range \( (t_{n-1}, t_{n}) \).
We therefore calculate execution time for each step as \( t_{c} - t_{0},\; t_{c} \in U(t_{n-1}, t_{n}) \), which on average works out to an adjustment of the observed time by 1.5 times the mean sampling rate of the step.
% This is also aggregated into an average at the block level.

% \begin{figure}
%   \centering
%   \includegraphics[width=.85\textwidth]{images/exec_time_diagram.pdf}
%   \caption{Visualization of the execution time of a step.}%
%   \label{fig:exectime:diagram}
% \end{figure}

@@

\begin{figure}[h]
  \centering
  \begin{subfigure}[t]{.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4a.eps}
    \caption{Structure of a step in a generic cognitive assistance application. 
    The assistant provides an instruction to the user and continuously samples the step state; inputs captured while the step is unfinished are silently ``discarded'' (i.e.\ they do not cause the generation of a new instruction).
    % by the assistant, as they do not contain relevant information.
    Once the user finishes performing the step, the next sample \emph{will} cause the generation of a new instruction.
    % which will subsequently be provided to the user.}%
    }
    \label{fig:cogassist:step}
  \end{subfigure}%
  % \medskip%
  \hfill%
  \begin{subfigure}[t]{.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4b.eps}
    \caption{In the experimental task, an additional variable segment of time is introduced immediately following the processing of the input frame in order to extend the perceived processing time of the input to a specific target delay.}%
    \label{fig:cogassist:step:delay}
  \end{subfigure}
  \medskip%
  \begin{subfigure}[t]{.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4c.eps}
    \caption{Structure of a block in the experimental task.}%
    \label{fig:cogassist:block}
  \end{subfigure}%
  % \medskip%
  \hfill%
  \begin{subfigure}[t]{.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4d.eps}
    \caption{Visualization of the execution time of a step.}%
    \label{fig:exectime:diagram}
  \end{subfigure}%
  % \medskip%
  \caption{Components of the cognitive assistance task.}
\end{figure}

@@

\begin{table}[h]
  \centering
  \caption{Means and standard deviations of normalized questionnaire scores.}\label{tab:qscores}
  \setlength{\tabcolsep}{0pt} 
  \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}clrr@{}}
    \toprule
    \textbf{Questionnaire} & \textbf{Metric} & \textbf{Mean} & \textbf{SD}\\
    \midrule
    \multirow{5}{*}{BFI} 
        & Agreeableness &               0.705 &                  0.136 \\
        & Conscientiousness &               0.562 &                  0.149 \\
        & Extroversion &               0.491 &                  0.180 \\
        & Neuroticism &               0.524 &                  0.230 \\
        & Openness &               0.677 &                  0.152 \\
    \cline{1-4}
    \multirow{4}{*}{ITQ} 
        & Focus &               0.626 &                  0.126 \\
        & Games &               0.463 &                  0.261 \\
        & Involvement &               0.568 &                  0.175 \\
        & Total &               0.540 &                  0.092 \\
    \bottomrule
  \end{tabular*}
\end{table}

@@

Both video feeds were captured at a rate of \num{24}~FPS in parallel processes to ensure a constant rate of capture.
Examples of the captured frames can be seen in \cref{fig:sampleframes}.
The video feeds were not used for the present study, but may be utilized in future analysis.

\begin{figure}[h]
  \centering
  \begin{subfigure}[t]{.35\textwidth}
    \centering
    \includegraphics[height=140pt]{publications/2021ImpactDelayedResponse/Fig5a.eps}
    \caption{Frame from the face recording of a random participant, clearly showing the locations of the EEG electrodes.}
  \end{subfigure}%
  % \par\bigskip
  % \hspace{.04\textwidth}%
  \hfill%
  \begin{subfigure}[t]{.60\textwidth}
    \centering
    \includegraphics[height=140pt]{publications/2021ImpactDelayedResponse/Fig5b.eps}
    \caption{Frame from the board recording of the same participant.}
  \end{subfigure}
  \caption{Sample frames captured during the experiment.}\label{fig:sampleframes}
\end{figure}

\section{Results}\label{sec:results}

@@

We confirmed this effect through an \emph{analysis of variance} (ANOVA~\cite{FUJIKOSHI1993ANOVA}) with factors of block length and delay, and found significant main effects of both factors and the interaction, shown in \cref{tab:anova:exectime}.
% An ANOVA test uses the \emph{F-test} statistic, which is a ratio of the variability in the data introduced by experimental manipulations (and their interactions) to the variability in the data attributable to randomness.
% The degrees of freedom represent the number of observations going into each of these variability measures.
% As we are using within-subject designs, the variability in the data due to randomness is estimated by the differences among subjects with respect to the effects of the experimental variables.
% The partial-\( \eta^2 \) (\(\eta_{p}^{2}\)) statistic is a measure of effect size, corresponding to the proportion of variance explained by the effect after excluding variance from others, and \( p \)-value, a measure of the probability of the effect under the null hypothesis.
% Conventionally \( p < .05 \) is the criterion for significance.
% The current length \( \times \) delay ANOVA found significant main effects of both factors and the interaction, shown in \cref{tab:anova:exectime}.

@@

\begin{table}[h]
  \centering
  \caption{Significant effects on per-step execution time from ANOVA on factors delay and block length.}\label{tab:anova:exectime}
  \setlength{\tabcolsep}{0pt}%
  \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}lrrr@{}}
    \toprule
    \textbf{Factor} & \textbf{F-test} & \( p \) & \( \eta^{2}_{p} \) \\
    \midrule
    length         &    \(F(2, 78) = 9.59\) &   \(< 0.001\) &           \(0.20\) \\
    delay          &  \(F(6, 234) = 15.52\) &  \(< 0.0001\) &           \(0.28\) \\
    length \(\times\) delay &  \(F(12, 468) = 3.84\) &  \(< 0.0001\) &           \(0.09\) \\
    \bottomrule
  \end{tabular*}
\end{table}

Further analysis focused on the progressive effect on per-step execution time as additional steps occurred at a constant delay.
For this purpose, the steps within a block were aggregated over sequences of 4, constituting a \emph{slice}.
Note that the first slice within a block is procedurally identical for all block lengths, in the sense that a participant currently performing a step in the first slice of a block has no way of predicting if the block ends after step 4 or not.
The same logic can be applied to slice 2 for blocks of length 8 and 12.
Accordingly for each participant, slice 1 data (steps 1--4) were pooled over all 3 block lengths, slice 2 (steps 5--8) were combined for block lengths 8 and 12, and slice 3 comprised the last four steps in blocks of length 12.
An ANOVA on slice number (1--3) and delay (7 values) yielded effects of slice number, delay, and the interaction, detailed in \cref{tab:anova:exectime:slice}.
As shown in \cref{fig:exectime:delay:slice}, blocks with shorter delays showed a trend for the execution time to progressively decline over the course of the block (i.e., by slice number), indicating that the participant accommodated to the feedback pace with more efficiently timed responses.
% We refer to the measure at slice 3 as a percentage of the measure at slice 1 as the ``change factor''.
With the longest delays, where the execution time per step was longest, the slow-down persisted; that is, the system response time hindered the participant's execution throughout the course of the block.

\begin{table}[h]
  \centering
  \caption{Significant effects on per-step execution time from ANOVA on factors block slice and delay.}\label{tab:anova:exectime:slice}
  \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
  \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}lrrr@{}}
    \toprule
    \textbf{Factor} & \textbf{F-test} & \( p \) & \( \eta^{2}_{p} \) \\
    \midrule
    slice         &   \( F(2, 78) = 88.79 \) &  \( < 0.0001 \) &         \( 0.69 \) \\
    delay         &  \( F(6, 234) = 14.13 \) &  \( < 0.0001 \) &         \( 0.27 \) \\
    slice \( \times \) delay &  \( F(12, 468) = 2.49 \) &    \( < 0.01 \) &         \( 0.06 \) \\
    \bottomrule
  \end{tabular*}%
\end{table}

\begin{figure}[h]
    \centering
    \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig7.eps}
    \caption{Mean per-step execution time vs.\ delay, by step slice.
    Error bars indicate S.E.M.}
    \label{fig:exectime:delay:slice}
\end{figure}

The data also allowed us to perform sub-analyses to assess the effects of carryover from one delay to another.
Specifically, we measured the per-step execution time for the first four steps of a block when participants transferred from a relatively long delay (\SIrange{2.175}{3.0}{\second}) versus a short delay (\SIrange{0}{1.65}{\second}).
Note that use of the first four steps controls for block length.
We performed analyses where participants transitioned from either a short- or long-delay block into a
\begin{enumerate*}[label=(\roman*), before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
  \item no delay block (36 subjects)
  \item \SI{1.65}{\second} delay block (40 subjects)
  \item \SI{2.7}{\second} delay block (40 subjects)
  \item \SI{3.0}{\second} delay block (37 subjects).
\end{enumerate*}
The destination delays were chosen so as to maximize the number of participants which contributed samples to the analyses, the results of which are pictured in \cref{fig:exectime:transition}.
We found that transitions from long-delay blocks carried over to significantly increase the per-step execution time of initial steps in the destination block for three of the four transitioned-into delays that were assessed (\SIlist{0;2.7;3.0}{\second}, all \( p < .025\)).

\begin{figure}[h]
  \centering
  \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig8.eps}
  \caption{Per-step execution time across the first four steps after a block transition from block \( B_{k-1} \) to \( B_k \). Error bars indicate S.E.M.}\label{fig:exectime:transition}%
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig10.eps}
  \caption{Movement score vs.\ delay, per block slice. Error bars indicate S.E.M.}\label{fig:acc:delayslice}
\end{figure}

>>>>

\cref{item:pervasive} implies use of lightweight and low-power devices, preferably a wearable device that frees both hands for work.
\cref{item:interaction}, on the other hand, suggests a level of context sensitivity and proactivity that can only be provided by real-time analysis of sensor inputs such as video and audio feeds.
The compute-intensive processing suggested by \cref{paper:olguinmunoz2021impact:item:interaction} cannot be met by the lightweight wearable devices suggested by \cref{item:pervasive}.
Only by offloading computation from a wearable device to cloud-based or edge-based infrastructure can this circle be squared.
However, offloading implies an extended end-to-end pipeline with many potential sources of queueing, transmission, and processing delays.
\cref{item:lowlatency} therefore emerges as a key concern, requiring deep understanding of the impact of end-to-end delays on \gls{WCA} users.

@@

The structure of this paper is as follows.
We describe the existing body of research around time perception and the effects of delay on human performance in \cref{paper:olguinmunoz2021impact:sec:background}.
\cref{sec:experimentaldesign} presents the experimental design, measures, and specific protocol.
Then, in \cref{sec:results} we detail the results of our experiment.
Implications for further modeling of the effects of delay are presented in \cref{paper:olguinmunoz2021impact:sec:discussion} before finally concluding the paper in \cref{sec:conclusion}.

\section{Background and Related Work}\label{paper:olguinmunoz2021impact:sec:background}

@@

\subsection{Time perception in computing systems}\label{paper:olguinmunoz2021impact:ssec:timeperception}

@@

\subsection{Mechanisms Relating Delay to Human Performance}\label{paper:olguinmunoz2021impact:ssec:potentialmechs}

@@

\section{Experimental Design}\label{paper:olguinmunoz2021impact:sec:experimentaldesign}

The core elements of our experiment are shown in \cref{paper:olguinmunoz2021impact:fig:experimentaltestbed}:
\begin{itemize}
    \item Subjects interact with a \gls{WCA} while wearing an array of biometric sensors.
    \item The \emph{responsiveness} of the application, i.e.\ the interval of time between an input being provided to the system and the associated output returned to the user, is manipulated in real time. 
    The effects of these manipulations on the subjects are recorded and subsequently analyzed.    
\end{itemize}

\begin{figure}[h]
  \centering
  \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig1.eps}
  \caption{Experimental test-bed.
  Participants interact with the cognitive assistant through task-related inputs and outputs --- in practice, these correspond to the video feed captured by the assistant and the instructions provided by it.
  The assistant itself has been instrumented with a data collection layer, which collects and processes experiment-related data such as biometric signals from the participants (these are merely processed here and do not form part of the inputs to the cognitive assistant as such, however), and a delay buffer, which introduces controlled delays in the transit of information from the core processing component.}%
  \label{paper:olguinmunoz2021impact:fig:experimentaltestbed}
\end{figure}

@@

% In more formal terms, we can provide definitions for \emph{task}, \emph{subtask}, and \emph{step} in such an application as:
In the following, we will provide some formal definitions relating to the \gls{WCA} application.
% \theoremstyle{definition}
% \begin{definition}[Task and subtask]
A \emph{task} will be understood as a finite sequence of instructions to be performed in order.
\emph{Step} will refer to a specific action to be performed by the user, described by a single instruction; e.g. \emph{put a green \(1 \times 3\) brick on the board} (see \cref{paper:olguinmunoz2021impact:fig:task:steps}).
% We will refer to the interval of time delimited by two consecutive instructions (see \cref{fig:task:steps,fig:cogassist:step}), characterized by the actions of both the user and the assistant.
A step starts with an instruction being given to the user, and ends when the assistant detects that the user has finished performing it.
At that point, the assistant either provides a new instruction to the user, or the task ends.
  % See for instance \cref{fig:task:steps}, which pictures the task of assembling a simple LEGO model, with each subtask corresponding to the addition of a specific LEGO piece to the current model.  
% \end{definition}

\begin{figure*}[h]
  \centering
  \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig2.eps}
  \caption{Example of a Cognitive Assistance task and its component steps.}\label{paper:olguinmunoz2021impact:fig:task:steps}
\end{figure*}

@@

\begin{figure*}[h]
  \centering
  % \input{plots/lego_hierarchy_diagram.tex}
  \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig3.eps}
  \caption{Hierarchical cognitive structure of a step in the LEGO task.}
  \label{paper:olguinmunoz2021impact:fig:lego:hierarchical}
\end{figure*}

@@

For the experimental LEGO task, a key modification was made to the structure of the steps.
After each the processing of each input is completed, the result is withheld for a variable period of time until a specific target \emph{delay} is reached, as illustrated in \cref{paper:olguinmunoz2021impact:fig:cogassist:step,fig:cogassist:step:delay}.
The length of this delay is one of two independent variables we manipulated for the experiment.
Seven levels of delay were used --- no added delay (which we will also refer to as \SI{0}{\second} delay), \SIlist{0.6;1.125;1.65;2.175;2.7;3.0}{\second} --- chosen based on the latency bounds found previously by Chen et al.~\cite{Chen:AnEmpiricalStudyOfLatency}.
% A value of  \SI{600}{\milli\second} was identified as the bound where users start noticing delays in the assistant.
% Conversely, \SI{2.7}{\second} was identified as the upper bound on delays after which the application is considered to be in such a degraded state it is basically ``unusable''.
Our selection of delays is centered around the range of delays where latency is noticeable to users, but the application remains in a ``usable'' state.
% while including one delay value in the unnoticeable range and one completely in the ``unusable'' range.

In order to study the effects of a delay applied across multiple steps we implement an experimental design component called a \emph{block}, corresponding to a sequence of consecutive steps within a task subject to the same delay; see \cref{paper:olguinmunoz2021impact:fig:cogassist:block}.
% \begin{definition}[Block]
%   A sequence of consecutive steps within a task subject to the same delay; see \cref{paper:olguinmunoz2021impact:fig:cogassist:block}.
The \emph{length} of blocks is the second of the two independent variables manipulated in our experiment.
We used values of \numlist{4;8;12} steps, chosen as representative of the number of steps in tasks in the base LEGO Assistant application.
Additionally, we define the \emph{duration} of a block, as the time elapsed between the start timestamp of the first step in the block and the end timestamp of the final step in the block.
% ; e.g.\ for \cref{fig:cogassist:block}, the duration of the pictured block would be \( t_{n+k} - t_{n} \).
% \end{definition}

@@

For an arbitrary step, we define the sequence \( \{ t_{0}, t_{1}, \ldots, t_{n} \} \) as the sequence of timestamps corresponding to the sampling instants during the step (see \cref{paper:olguinmunoz2021impact:fig:exectime:diagram}).
\( t_{0} \) corresponds to the instant when the instruction for the step is given to the user (and the first sample is taken), and \( t_{n} \) to the timestamp of the last sample before a new instruction is given.
If we define \( t_{c} \) as the instant marking when the user presented the finished step to the system, it must follow that \( t_{n-1} < t_{c} < t_{n} \).
Due to the discrete sampling of the task state, there remains some imprecision in the estimate of execution time relative to \( t_{c} \).
However, this introduces no bias in the results, as we can assume that \( t_{c} \) is uniformly distributed in the range \( (t_{n-1}, t_{n}) \).
We therefore calculate execution time for each step as \( t_{c} - t_{0},\; t_{c} \in U(t_{n-1}, t_{n}) \), which on average works out to an adjustment of the observed time by 1.5 times the mean sampling rate of the step.
% This is also aggregated into an average at the block level.

% \begin{figure}
%   \centering
%   \includegraphics[width=.85\textwidth]{images/exec_time_diagram.pdf}
%   \caption{Visualization of the execution time of a step.}%
%   \label{paper:olguinmunoz2021impact:fig:exectime:diagram}
% \end{figure}

@@

\begin{figure}[h]
  \centering
  \begin{subfigure}[t]{.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4a.eps}
    \caption{Structure of a step in a generic cognitive assistance application. 
    The assistant provides an instruction to the user and continuously samples the step state; inputs captured while the step is unfinished are silently ``discarded'' (i.e.\ they do not cause the generation of a new instruction).
    % by the assistant, as they do not contain relevant information.
    Once the user finishes performing the step, the next sample \emph{will} cause the generation of a new instruction.
    % which will subsequently be provided to the user.}%
    }
    \label{paper:olguinmunoz2021impact:fig:cogassist:step}
  \end{subfigure}%
  % \medskip%
  \hfill%
  \begin{subfigure}[t]{.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4b.eps}
    \caption{In the experimental task, an additional variable segment of time is introduced immediately following the processing of the input frame in order to extend the perceived processing time of the input to a specific target delay.}%
    \label{paper:olguinmunoz2021impact:fig:cogassist:step:delay}
  \end{subfigure}
  \medskip%
  \begin{subfigure}[t]{.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4c.eps}
    \caption{Structure of a block in the experimental task.}%
    \label{paper:olguinmunoz2021impact:fig:cogassist:block}
  \end{subfigure}%
  % \medskip%
  \hfill%
  \begin{subfigure}[t]{.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4d.eps}
    \caption{Visualization of the execution time of a step.}%
    \label{paper:olguinmunoz2021impact:fig:exectime:diagram}
  \end{subfigure}%
  % \medskip%
  \caption{Components of the cognitive assistance task.}
\end{figure}

@@

\begin{table}[h]
  \centering
  \caption{Means and standard deviations of normalized questionnaire scores.}\label{paper:olguinmunoz2021impact:tab:qscores}
  \setlength{\tabcolsep}{0pt} 
  \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}clrr@{}}
    \toprule
    \textbf{Questionnaire} & \textbf{Metric} & \textbf{Mean} & \textbf{SD}\\
    \midrule
    \multirow{5}{*}{BFI} 
        & Agreeableness &               0.705 &                  0.136 \\
        & Conscientiousness &               0.562 &                  0.149 \\
        & Extroversion &               0.491 &                  0.180 \\
        & Neuroticism &               0.524 &                  0.230 \\
        & Openness &               0.677 &                  0.152 \\
    \cline{1-4}
    \multirow{4}{*}{ITQ} 
        & Focus &               0.626 &                  0.126 \\
        & Games &               0.463 &                  0.261 \\
        & Involvement &               0.568 &                  0.175 \\
        & Total &               0.540 &                  0.092 \\
    \bottomrule
  \end{tabular*}
\end{table}

@@

Both video feeds were captured at a rate of \num{24}~FPS in parallel processes to ensure a constant rate of capture.
Examples of the captured frames can be seen in \cref{paper:olguinmunoz2021impact:fig:sampleframes}.
The video feeds were not used for the present study, but may be utilized in future analysis.

\begin{figure}[h]
  \centering
  \begin{subfigure}[t]{.35\textwidth}
    \centering
    \includegraphics[height=140pt]{publications/2021ImpactDelayedResponse/Fig5a.eps}
    \caption{Frame from the face recording of a random participant, clearly showing the locations of the EEG electrodes.}
  \end{subfigure}%
  % \par\bigskip
  % \hspace{.04\textwidth}%
  \hfill%
  \begin{subfigure}[t]{.60\textwidth}
    \centering
    \includegraphics[height=140pt]{publications/2021ImpactDelayedResponse/Fig5b.eps}
    \caption{Frame from the board recording of the same participant.}
  \end{subfigure}
  \caption{Sample frames captured during the experiment.}\label{paper:olguinmunoz2021impact:fig:sampleframes}
\end{figure}

\section{Results}\label{paper:olguinmunoz2021impact:sec:results}

@@

We confirmed this effect through an \emph{analysis of variance} (ANOVA~\cite{FUJIKOSHI1993ANOVA}) with factors of block length and delay, and found significant main effects of both factors and the interaction, shown in \cref{paper:olguinmunoz2021impact:tab:anova:exectime}.
% An ANOVA test uses the \emph{F-test} statistic, which is a ratio of the variability in the data introduced by experimental manipulations (and their interactions) to the variability in the data attributable to randomness.
% The degrees of freedom represent the number of observations going into each of these variability measures.
% As we are using within-subject designs, the variability in the data due to randomness is estimated by the differences among subjects with respect to the effects of the experimental variables.
% The partial-\( \eta^2 \) (\(\eta_{p}^{2}\)) statistic is a measure of effect size, corresponding to the proportion of variance explained by the effect after excluding variance from others, and \( p \)-value, a measure of the probability of the effect under the null hypothesis.
% Conventionally \( p < .05 \) is the criterion for significance.
% The current length \( \times \) delay ANOVA found significant main effects of both factors and the interaction, shown in \cref{paper:olguinmunoz2021impact:tab:anova:exectime}.

@@

\begin{table}[h]
  \centering
  \caption{Significant effects on per-step execution time from ANOVA on factors delay and block length.}\label{paper:olguinmunoz2021impact:tab:anova:exectime}
  \setlength{\tabcolsep}{0pt}%
  \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}lrrr@{}}
    \toprule
    \textbf{Factor} & \textbf{F-test} & \( p \) & \( \eta^{2}_{p} \) \\
    \midrule
    length         &    \(F(2, 78) = 9.59\) &   \(< 0.001\) &           \(0.20\) \\
    delay          &  \(F(6, 234) = 15.52\) &  \(< 0.0001\) &           \(0.28\) \\
    length \(\times\) delay &  \(F(12, 468) = 3.84\) &  \(< 0.0001\) &           \(0.09\) \\
    \bottomrule
  \end{tabular*}
\end{table}

Further analysis focused on the progressive effect on per-step execution time as additional steps occurred at a constant delay.
For this purpose, the steps within a block were aggregated over sequences of 4, constituting a \emph{slice}.
Note that the first slice within a block is procedurally identical for all block lengths, in the sense that a participant currently performing a step in the first slice of a block has no way of predicting if the block ends after step 4 or not.
The same logic can be applied to slice 2 for blocks of length 8 and 12.
Accordingly for each participant, slice 1 data (steps 1--4) were pooled over all 3 block lengths, slice 2 (steps 5--8) were combined for block lengths 8 and 12, and slice 3 comprised the last four steps in blocks of length 12.
An ANOVA on slice number (1--3) and delay (7 values) yielded effects of slice number, delay, and the interaction, detailed in \cref{paper:olguinmunoz2021impact:tab:anova:exectime:slice}.
As shown in \cref{fig:exectime:delay:slice}, blocks with shorter delays showed a trend for the execution time to progressively decline over the course of the block (i.e., by slice number), indicating that the participant accommodated to the feedback pace with more efficiently timed responses.
% We refer to the measure at slice 3 as a percentage of the measure at slice 1 as the ``change factor''.
With the longest delays, where the execution time per step was longest, the slow-down persisted; that is, the system response time hindered the participant's execution throughout the course of the block.

\begin{table}[h]
  \centering
  \caption{Significant effects on per-step execution time from ANOVA on factors block slice and delay.}\label{paper:olguinmunoz2021impact:tab:anova:exectime:slice}
  \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
  \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}lrrr@{}}
    \toprule
    \textbf{Factor} & \textbf{F-test} & \( p \) & \( \eta^{2}_{p} \) \\
    \midrule
    slice         &   \( F(2, 78) = 88.79 \) &  \( < 0.0001 \) &         \( 0.69 \) \\
    delay         &  \( F(6, 234) = 14.13 \) &  \( < 0.0001 \) &         \( 0.27 \) \\
    slice \( \times \) delay &  \( F(12, 468) = 2.49 \) &    \( < 0.01 \) &         \( 0.06 \) \\
    \bottomrule
  \end{tabular*}%
\end{table}

\begin{figure}[h]
    \centering
    \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig7.eps}
    \caption{Mean per-step execution time vs.\ delay, by step slice.
    Error bars indicate S.E.M.}
    \label{paper:olguinmunoz2021impact:fig:exectime:delay:slice}
\end{figure}

The data also allowed us to perform sub-analyses to assess the effects of carryover from one delay to another.
Specifically, we measured the per-step execution time for the first four steps of a block when participants transferred from a relatively long delay (\SIrange{2.175}{3.0}{\second}) versus a short delay (\SIrange{0}{1.65}{\second}).
Note that use of the first four steps controls for block length.
We performed analyses where participants transitioned from either a short- or long-delay block into a
\begin{enumerate*}[label=(\roman*), before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
  \item no delay block (36 subjects)
  \item \SI{1.65}{\second} delay block (40 subjects)
  \item \SI{2.7}{\second} delay block (40 subjects)
  \item \SI{3.0}{\second} delay block (37 subjects).
\end{enumerate*}
The destination delays were chosen so as to maximize the number of participants which contributed samples to the analyses, the results of which are pictured in \cref{paper:olguinmunoz2021impact:fig:exectime:transition}.
We found that transitions from long-delay blocks carried over to significantly increase the per-step execution time of initial steps in the destination block for three of the four transitioned-into delays that were assessed (\SIlist{0;2.7;3.0}{\second}, all \( p < .025\)).

\begin{figure}[h]
  \centering
  \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig8.eps}
  \caption{Per-step execution time across the first four steps after a block transition from block \( B_{k-1} \) to \( B_k \). Error bars indicate S.E.M.}\label{paper:olguinmunoz2021impact:fig:exectime:transition}%
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig10.eps}
  \caption{Movement score vs.\ delay, per block slice. Error bars indicate S.E.M.}\label{paper:olguinmunoz2021impact:fig:acc:delayslice}
\end{figure}
